Type Classes for Mathematics in Type Theory

### 2010-12-15 papers/mscs/mscs.tex

% preliminary choices:
%   don't discuss difference between maximally-inserted and non-maximally-inserted implicit arguments.

% todo:
  % cite that other type class based development in the context of the alternative bundling schemes.

@@

\begin{document}
\title{Type Classes for Mathematics in Type Theory}
\author{Bas Spitters \and Eelis van der Weegen} % For MSCS \\ Radboud University Nijmegen}
\institute{Radboud University Nijmegen}
\date{today}
\maketitle
\begin{abstract}
The introduction of first-class type classes in the Coq system calls for re-examination of the basic interfaces used for mathematical formalization in type theory.
% especially since existing solutions fall short from a proof engineering perspective.
We present a new set of type classes for mathematics and take full advantage of their unique features to make practical a particularly flexible approach formerly thought infeasible. Thus, we address both traditional proof engineering challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a minimum.

@@

% ideally should be more or less readable by Haskell person. that would be neat. so that's what the preliminaries should prepare for.
In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transferred
to any type theory based proof assistant supporting type classes such as
Matita~\cite{asperti2007user}.

@@

 %Having derived some principles concerning bundling, we then proceed to show (in section \ref{managing}) how type classes may be employed to adopt these principles without compromising ease of use. Next, in section \ref{hierarchy}, we introduce a number of additional conventions that let us complete a consistent idiom for interfaces to algebraic structures. With this in place, we continue with a discussion of

@@

% ideally, this should make the paper readable for Haskell people

@@

Propositions in Coq are types~\cite{CMCP,ITT}, which themselves have types called \emph{sorts}. Coq features a distinguished sort called \lstinline|Prop| that one may choose to use as the sort for types representing propositions. The distinguishing feature of the \lstinline|Prop| sort is that terms of non-\lstinline|Prop| type may not depend on the values of inhabitants of \lstinline|Prop| types (that is, proof terms).
% As originally conceived~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
% them. Often we do \emph{not} want functions to depend on proofs. In Coq, this is the paradigmatic way of working. We are forbidden to extract information from proofs of \lstinline|Prop|s. Especially with the aid of the Program machinery, this allows us to flexibly work with simply typed program and their Hoare style correctness proofs.
This regime of discrimination establishes a weak form of proof irrelevance, in that changing a proof can never affect the result of value computations. On a very practical level, this lets Coq safely erase all \lstinline|Prop| components when extracting certified programs to OCaml or Haskell.

% A stronger notion of proof irrelevance in which any two inhabitants of a \lstinline|Prop| type are made \emph{convertible} is planned for a future Coq version.
  % commented out because not really relevant. only re-add with explanation of relevance.

% todo: process?:
%One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
%construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. In Coq,
%the [ ]-types are isomorphic to the type of sort \lstinline|Prop|. However,
%there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
%theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
%irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
%of Propositions as types seems necessary.
% Making Harrop formulas proof irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. An orthogonal and further reaching proposal is to replace the Coq type theory with the Calculus of implicit constructions with implicit Σ-types~\cite{miquel2001implicit,barras2008implicit,Bernardo}.

@@

% However, we would also like to use Coq as a dependently typed programming language. This requires a hybrid approach, were we put informative propositions in \lstinline|Type| and non-informative ones in \lstinline|Prop|. 
% Eelis: I don't see the logic here.

@@

% \section{The Type-Classified Algebraic Hierarchy}\label{classes}
% We represent each structure in the algebraic hierarchy as a type class. This immediately leads to the familiar question of which components of the structure should become parameters of the class, and which should become fields. By far the most important design choice in our development is the decision to turn all \emph{structural} components (i.e. carriers, relations, and operations) into parameters, keeping only \emph{properties} as fields. Type classes defined in this way are essentially predicates with automatically resolved proofs.
% 
% Conventional wisdom warns that while this approach is theoretically very flexible, one risks extreme inconvenience both in having to declare and pass around all these structural components all the time, as well as in losing notations (because we no longer project named operations out of records). These legitimate concerns will be addressed in Sections~\ref{bundle}-\ref{classes}. Section~\ref{univ} abstracts to universal
% algebra, we develop the basic theory to the point of the first homomorphism theorem.
% Section~\ref{interfaces} provides interfaces to usual data structures by providing their universal
% properties. To allow symbolic manipulation of semantic objects one needs a quote function. Usually,
% this function is implemented in Coq's tactics language. A more convenient way is to use type class
% unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparison
% between type classes and canonical structures.

@@

% hm, i used to have paragraphs here about how it would be unclear how this approach could support structure inference and structure inheritance, but those are actually pretty straightforward using ordinary proof search.

@@

% While not observable in this example, when combined with other specialized facilities in Coq (namely coercions and/or canonical structures), this bundled representation also lets one address the problems of structure inference inheritance to some extent.

@@

%For completeness, we mention an alternative record-based approach known as ``packed classes'', as used in the ssreflect library. In the packed classes idiom, each algebraic structure is represented by not one but a trinity of interdependent records along with packing/unpacking functions. These are then stacked in an intricate manner designed to reach a particular hybrid packing balance that, combined with the use of canonical structures and coercions, accomplishes some of the goals above.

%We have found it hard to assess the general merits of this approach, partly because of its complexity, and partly because the only actual usage example (the hierarchy in the ssreflect library) is not only wedded to a very specialized view on equality and decidability that does not support setoid relations from the outset, but also diverges significantly from standard mathematical dichotomy, making it hard to recognize familiar structures and presentation. The reasons for this (as we understand them) are pragmatic and twofold.

%First, the ssreflect library was originally developed strictly to support very practical work in a specialized domain of mathematics involving decidable and mostly finite structures. Second, at the time of ssreflect's initial development, support for setoid rewriting in Coq was not nearly as stable and refined as it is today, which made ``Leibniz equality at all costs'' a more reasonable strategy at the time than it might be considered today.

%Work to extend the ssreflect framework with techniques to allow representation of a less restricted class of mathematical structures has only recently begun and is still ongoing, though there are currently no plans to support undecidable structures.

%Because in our development we will be working with computable reals, equality on which can never be either Leibniz or decidable, we conclude that the ssreflect library is not usable for us.

@@

% Another benefit of our approach is that it works very smoothly when these records are in fact type
% classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
% structure record/class, to express that some collection of relations and
% operations possess the structure as a type class constraint seems to require manifest fields; see
% Section~\ref{manifest}. With
% our unbundled approach, on the other hand, such type class constraints are easily expressed without
% any extensions as type inhabitation queries
% % (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
% of types living in Prop, so that we don't even care what instances are found (thus embracing proof
% irrelevance).

% Leaving the equivalence proofs bundled means that when one has a proof that a structure
% is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
% equivalence. Fortunately, this redundancy is entirely harmless since our terms
% never refer (directly or indirectly) to proofs, which is the case when the relations and operations
% they are composed from are left unbundled. Here, too, we enjoy the warm embrace of proof
% irrelevance.
% 
% From the discussion so far, we derive some principles:
% \begin{enumerate}
%  \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
% (i.e. carriers, relations, and operations) as parameters, and nothing else;
%  \item (as a consequence) these concepts should be defined as predicates (which includes records and
% classes living in Prop), and their proofs may be kept opaque;
%  \item (as a second consequence) algebraic expressions need/should never refer (directly or
% indirectly) to proofs about the algebraic structure.\footnote{The reciprocal in a field
% famously does refer to a proof that the element is non-zero, but not to a proof that the operations
% indeed form a field.}
% \end{enumerate}
% % \setcounter{enumi}{3}
% 
% The principles listed so far are not particularly controversial, in fact this
% approach ensures maximum flexibility~\cite{Hints}. However,
% there is a common perception that unbundling carriers, relations, and operations this way invariably
% leads either to unmanageably big algebraic expressions (involving endless projections and/or implicit
% arguments), or to unmanageably long and tedious enumeration of carriers/relations/operations, or to
% problems establishing canonical names for operations (since we no longer project them out of
% structure instances all the time).
% 
% In the next sections, we show how systematic use of type classes (and their support infrastructure
% in Coq), combined with the use of a special form of type class we call an `operational type class'
% (as opposed to a `structural' type class we might use in place of the records shown thus far), lets
% us avoid these problems.

@@

% We note that operational type classes allow us to avoid Coq's notation \lstinline|scope| mechanism.
% hm, i don't think this is true. we don't need scopes for different representations' use of operators because we properly recognize their commonality, but that can be done without operational type classes as well.

% TODO: get this in:
% Although the expression now \emph{looks} like it is bound to some
% specific semigroup structure (which with traditional approaches would imply projections from bundles
% and/or reference to proofs), it is really only referring near-directly to the actual operations
% involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
% may use in the proof. This lack of projections keeps our terms small and independent, and keeps
% rewriting simple and sane; see Section~\ref{canonical}.

@@

All of \lstinline|Equiv|, \lstinline|RingPlus|, \lstinline|RingMult|, \lstinline|RingZero|, and \lstinline|RingOne| are operational (single-field) classes, with bound notations \lstinline|=|, \lstinline|+|, \lstinline|*|, \lstinline|0|, and \lstinline|1|, respectively.
\begin{wrapfigure}{r}{.4\textwidth}
    \includegraphics[scale=.8]{hierarchy.pdf}
%   \caption{}
\end{wrapfigure}
Let us briefly highlight some additional aspects of this style of structure definition in more detail.

@@

% note: we don't mention "super classes" because by Matthieu's terminology only parameters-of-class-type are "super classes", and for us that means only the operational classes, which we certainly don't think of as being part of the hierarchy. Matthieu calls fields of class type declared with ":>" substructures.

@@

% todo: problem with ambiguous projection paths: quoting will not notice convertibility.

@@

% rather strict separation of operations and proofs of properties about them (resolved by instance resolution) provides a very pleasing environment to work in, without any need for hacks.
% terms composed of algebraic operations and relations 
% separation of , this kind of redundance is never a problem for us. 
% it is /never/ a problem to know a fact twice.
% multiple proofs of equality, and even of semigroup-ness (since both Monoid), but that's simply not a problem, since none of our terms ever refers to such proofs. We have found that this rather strict separation of operations and proofs of properties about them (resolved by instance resolution) provides a very pleasing environment to work in, without any need for hacks.

@@

% : with predicate classes we can essentially pose queries of the form ``do I already know that proposition P holds?'', whereas canonical structures
% ``do I already have a composite structure X of which Y is a specific component?
% 
% arbitrary predicates, based on associating particular

% Having argued that the \emph{all-structure-as-parameters} approach \emph{can} be made practical, we enumerate some of the benefits that make it worthwhile.
% 
% First, multiple inheritance becomes trivial: \lstinline|SemiRing| inherits two \lstinline|Monoid| structures on the same carrier and setoid relation, using ordinary named arguments to achieve ``manifest fields''; see section~\ref{manifest}.

% Third, since our structural type classes are mere predicates, overlap between their instances is a non-issue. Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses: by simply declaring a \lstinline|SemiRing| class instance showing that a ring is a semiring, results about semirings immediately apply implicitly to any known ring, without us having to explicitly encode this relation in the hierarchy definition itself, and without needing any projection or translation of carriers or operations.
% 
% Fourth, \emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
% can find the equivalence relation in the context.\footnote{
% A similar style should be possible for, say, the \lstinline|Ring| tactic, instead of
% declaring the ring structure by a separate command, we would rely on type class resolution to find it.}
% We note that \lstinline|op| does not depend on the proof that \lstinline|e| is an equivalence. As explained above we use Coq's implicit quantification (\lstinline|`{}|) to avoid having to write all the parameters when \emph{stating} a theorem and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a theorem. Both features are new in Coq and stem from the type class implementation.
% 
% We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
% those of telescopes or packed classes. In our experience, this has been relatively
% harmless\footnote{However, Coq's data structure for contexts is not very efficient. Gonthier fears that this
% may be a bottleneck for huge developments. It seems that the data structure chosen
% in~\cite{asperti2009compact} will behave better.}%
% : most terms in the context are there to support canonical names. Bigger terms
% \emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
% proofs to allow rewriting. Moreover, it prohibits Opaque proofs and `proof irrelevance'. 2. The
% projection paths may not be canonical.
% 
% Coercion pullbacks~\cite{Hints} were introduced to address problems with multiple coercions paths,
% as in the definition of a semiring: a type with two monoid structures on it. We avoid some
% of these problems by explicitly specifying the fields. We emphasize that the semiring properties are
% automatically derived from the ring properties, although the properties of a semiring are not
% structurally included in the ring properties.
% 
% Let us now stop and think to what extent this approach suffers from all the problems commonly
% associated with it. In particular, let us imagine what happens to our terms and contexts when we
% want to talk about nested structures such as polynomials over the ring. For a concrete
% representation of the polynomials the context will just contain the context for an abstract ring.
% For abstract reasoning about polynomials the context will grow with abstract operations on
% polynomials. Consequently, the context will grow linearly, as opposed to exponentially.

@@

With these in place, our type class for categories follows the usual type-theoretical definition of a
category~\cite{saibi1995constructive}:
\begin{lstlisting}
  Class Category (O: Type) `{Arrows O} `{∀$$x y: O, Equiv (x ⟶ y)}
      `{CatId O} `{CatComp O}: Prop :=
    { arrow_equiv:> ∀ x y, Setoid (x ⟶ y)
    ; comp_proper:> ∀ x y z, Proper (equiv ==> equiv ==> equiv) comp
    ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
        c ◎$$ (b ◎$$ a) = (c ◎$$ b) ◎$$ a
    ; id_l `(a: x ⟶ y): cat_id ◎$$ a = a
    ; id_r `(a: x ⟶ y): a ◎$$ cat_id = a }.
\end{lstlisting}
This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
%Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?

@@

We introduce a convenient notation for the type of the computational content of a natural transformation between two functors:
\begin{lstlisting}
  Notation "F ⇛ G" := (∀ x, F x ⟶ G x).
\end{lstlisting}
Now assume the following context:
\begin{lstlisting}
  Context `{Category C} `{Category D}
    `{Functor (F: C → D)} `{Functor (G: D → C)}.
\end{lstlisting}
The naturality property is easy to write:
\begin{lstlisting}
  Class NaturalTransformation (η: F ⇛ G): Prop :=
    { naturaltrans_from: Functor F
    ; naturaltrans_to: Functor G
    ; natural: ∀ `(f: x ⟶ y), η$$ y ◎$$ fmap F f = fmap G f ◎$$ η$$ x }.
\end{lstlisting}
  % Todo: Having to add "$$" to get whitespace is awful. we can't add "\ " in η's literate-replacement though, because then we also get space before the colon in "η: t".

% We have defined the category of setoids, the dual of a category, products, and hence co-products.

% The usual size problems in the definition of the category of categories can be avoided by using universe polymorphism. 
% However, we need to avoid making \lstinline|Relation| a \lstinline|Definition|, since \lstinline|Definitions| are not (yet?) universe polymorphic.

@@

% \footnote{In Bishop's words~\cite[p.12]{Bishop/Bridges:1985}: The axiom of choice is used to extract
%elements from equivalence classes where they should never have been put in the first place.}

@@

The rationals are characterized as a decidable field with an injective ring morphism from a canonical implementation of the integers and a surjection of fractions of such integers:
\begin{lstlisting}
  Context `{Field A} `{∀$$x y: A, Decision (x = y)} {inj_inv}.
  Class Rationals: Prop :=
    { rationals_field:> Field A
    ; rationals_frac: Surjective
        (λ p => integers_to_ring (Z nat) A (fst p) *
          / integers_to_ring (Z nat) A (snd p)) (inv:=inj_inv)
    ; rationals_embed_ints: Injective (integers_to_ring (Z nat) A) }.
\end{lstlisting}
Here, \lstinline|Z| is an \lstinline|Integers| implementation paramerized by a \lstinline|Naturals| implementation, for which we just take \lstinline|nat|. The choice of \lstinline|Z nat| here is immaterial; we could have picked another, or even a generic, implementation of \lstinline|Integers|, but doing so would provide no benefit.
  % todo: hm, this Z is poorly named. it's really natpair_integers.Z
  % todo: say a few words about how we handle division by zero?

% todo: say a word about section variable/context discharging, because we use it everywhere.

@@

% Eelis: and when we do, we can write something like:
% Given a ring $R$, the $R$-algebra $R[X]$ of polynomials, is the free $R$-algebra on a set $X$.
% We provide two implementations of polynomials: the
% standard representation using lists of coefficients and the Bernstein representation. A Bernstein
% basis polynomial is one of the form:
% \[b_{\nu,n}(x) = {n \choose \nu} x^{\nu} \left( 1 - x \right)^{n - \nu}, \quad \nu = 0, \ldots, n.\]
% A Bernstein polynomial is a linear combination of these basic polynomials. Bernstein polynomials
% have been used for efficient computations inside the Coq system~\cite{ZumkellerPhD}.

% Eelis: TODO: process: At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category \lstinline|Cat| of categories, and finitary algebraic categories  defined by a theory which we instantiate to obtain the categories of monoids, semirings, and rings), products

%\subsection{Categorical universal algebra}

% We define the forgetful functor $U$ from the category of τ-algebras to the category of sets and its left-adjoint $F$: the term algebra consisting of closed terms, i.e.\ terms with the empty type as set of variables. This term algebra is the initial τ-algebra. Every adjunction defines a monad by composition of the functors $U$ and $F$. We will use the resulting expression monad in Section~\ref{quote} below.
% Eelis: oh?

%The product of two $\tau$-algebras is their categorical product.
% 
% \begin{comment}
% Given a sub\emph{set} of an algebra, we define the sub\emph{algebra} generated by it.
% An interesting fact to mention is the use of the following heterogeneous equality between
% elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
% \begin{lstlisting}
%   Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
%     match o with
%     | constant x => fun a b => `a == b
%     | function x y => fun a b => forall u, heq (a u) (b (`u))
%     end.
% \end{lstlisting}
% \end{comment}
% 
% We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
% of monoids. Given a (concrete) monoid, we can construct the corresponding $\tau_m$-algebra. 
% Conversely, given a $\tau_m$-algebra, we construct an instance of the \lstinline|Monoid| type class.
% %varieties/monoid.v
% \begin{lstlisting}
% Variable o: variety.Object theory.
% Global Instance: SemiGroupOp (o tt) := algebra_op theory mult.
% Global Instance: MonoidUnit (o tt) := algebra_op theory one.
% Global Instance from_object: Monoid (o tt).
% \end{lstlisting}
% 
% In the last line, instance resolution `automatically' finds the operation and the unit specified in
% the lines before.
% 
% This interplay between concrete algebraic structure and their expressions on the one hand, and models of equational theories on the other is occasionally a source of tension (in that translation in either direction is not yet fully automatic). However, it opens the door to the possibility of fully internalized implementations of generic tactics for algebraic manipulation, no longer requiring plugins. We come back to this when we describe automatic quotation of concrete expressions into universal algebra expressions; see Section~\ref{quote}.

@@

For ease of presentation we show only a proof of concept for a very concrete language. We are currently working to integrate this technique with our existing universal algebra infrastructure. In particular, the latter's term data type should be ideally suited to serve as a generic symbolic representation of terms in a wide class of algebras. This should let us implement the basic setup of the technique once and for all, so that quotation for new algebraic structures can be enabled with minimal effort.
  %  Preliminary results in this area (such as a simple monoid normalization tactic) suggest yada.

@@

Suppose now that we want to quote \lstinline|nat| expressions built from 1 and multiplication. To describe the relation we want the symbolic expression to have to the original expression, we first define how symbolic expressions evaluate to values (given a variable assignment):
%An expression is only meaningful in the context of a variable assignment:\footnote{Categorically: \lstinline|nat| is an \lstinline|Expr|-algebra and \lstinline|eval| is \lstinline|map vars| composed with the algebra map.}
\begin{lstlisting}
  Definition Value := nat.
  Definition Env V := V → Value.

@@

%Monads are trees with grafting~\cite{MonadsGrafting}: the tree monad
%$TX:=1+T^2X$, may be seen as the prototypical monad. 
% The \lstinline|Expr| monad, i.e.\ the free \lstinline|Expr|-algebra monad on the category of \lstinline|Type|s with extensional functions%\marginpar{Proof in classquote}
% , provides a bind operation which describes the arrows in the Kleisli category. Concretely,
% \begin{lstlisting}
% Fixpoint bind {V W} (f: V → Expr W) (e: Expr V): Expr W :=
%   match e with
%   | Zero => Zero
%   | Mult a b => Mult (bind f a) (bind f b)
%   | Var v => vs v
%   end.
% \end{lstlisting}
% Eelis: above commented out because we never actually use it, and it's really not pertinent to the discussion.

% We have shown that |Expr|, i.e.\ |bind| together with |Var|, is a monad on the category of Types
% with extensional functions between them~\marginpar{Really?}.

@@

%A similar method is used in Coq's notation mechanism.
% We recommend a change in
% implementation supporting an agda style precedence relation which is only restricted to be a
% directed acyclic graph~\cite{danielsson2009parsing}.

% We have used this technique to implement a tactic to normalize monoid expressions, and plan to extend yada yada.
%It would be interesting to use this technique to implement the \lstinline|congruence| tactic, which implements
%the congruence closure algorithm~\cite{corbineau2007deciding}. One would naturally obtain a tactic
%which moreover works on setoid equalities.

@@

  Class Extend := extend: ∀ {x y} `{SemiGroupOp y} `{MonoidUnit y},
    (x → y) → (seq x → y).
  Class Singleton := singleton: ∀ x, x → seq x.
\end{lstlisting}
With these, we can define the predicate class for sequences:
\begin{lstlisting}
  Class Sequence
   `{∀ a, MonoidUnit (seq a)} `{∀ a, SemiGroupOp (seq a)}
   `{∀ a, Equiv a → Equiv (seq a)} `{Singleton} `{Extend}: Prop := ...
\end{lstlisting}
%    { ... structural and propriety proofs sufficient to bundle up the categories and arrows ...
%    ; ... AltAdjunction stated using the bundled ... }.
On top of this interface we can build theory about typical sequence operations such as maps, folds, their relation to \lstinline|singleton| and \lstinline|extend|, et cetera. We can also generically define `big operators' for sums ($\sum$) and products ($\prod$) of sequences, and easily show properties like distributivity, all without ever mentioning cons-lists.

@@

% todo: not sure why we didn't have this problem with, say, nat/Naturals.

\begin{comment}
Canonical structures have been used to provide a uniform treatment~\cite{bertot2008canonical} of big
operators, like $\Pi,\Sigma, \max$. These operators extend a pair of a binary and a 0-ary operation
to an $n$-ary operation for any $n$. Categorically, one considers the algebra maps from non-empty
lists, lists, inhabited finite sets and finite sets to the carrier of a semigroup, monoid,
commutative semigroup, commutative monoid. Hence we want to reuse the libraries for lists etc.\ as much as possible.

Again, we can use type classes, instead of canonical structures, to deduce the relevant monoid operation:
\begin{lstlisting}
  Definition seq_sum
    `{Sequence A T} `{RingPlus A} `{z: RingZero A}: T $\to$ A
    := @seq_to_monoid A T _ A ring_plus z id.
  Eval compute in seq_sum [3; 2].
\end{lstlisting}
\end{comment}

\begin{comment}
\section{Subset types and proof irrelevance}\label{PI}
We have mentioned proof irrelevance a number of times before. It is an important theme in our
approach. The \lstinline|Program| machinery allows one to conveniently write programs with Hoare-style
pre- and post-conditions. I.e.\ functions $f: \{ A \mid P \} \to \{ B \mid Q \}$. Both side
conditions
are intended to be proof irrelevant, they are in \lstinline|Prop|. Presently, Coq does not support such subset
types, thus forcing the user to manually prove that \lstinline|f| does not depend on the proof of \lstinline|P|.
The addition of proof irrelevance and subset types to Coq is pursued by
Barras~\cite{Barras:subset,Werner08}. 
% Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
% \ldots.

Our approach may be intuitively explained by `telescopic' subset types. We recall the use of telescopes in record types~\cite{pollack2000dependently}. Given a telescope\[
T=[x_1:A_1][x_2:A_2(x_1)]\cdots[x_k:A_k(x_1,\ldots,x_k)].
\]
There is an inductively defined type $\Sigma T$, with a single constructor, given by the following 
formation and introduction rule in pseudo-code for $k=2$:
\begin{align*}
 A_1 :& Type\\
 A_2 :& \Pi A_1,Type\\
\cline{1-2}
\Sigma T :& Type
\end{align*}

\begin{align*}
 \Sigma T:Type\quad a_1 &:A_1 \quad a_2 : A_2(a_1)\\
\cline{1-3}
(a_1,a_2) &: \Sigma T
\end{align*}

For our methodology to work we have to be able to transform the occurring record into a telescopic
subset $\{ A \mid  B\}$. We moreover conjecture that the terms in \lstinline|B| only depend on the
variables in \lstinline|A|, but not in \lstinline|B| itself. However, we do not need this.

We are naturally lead to the following methodology: Let $\phi$ be a statement in type
theory, i.e.\ an alternation of $\Pi$s and $\Sigma$s. By the type theoretic axiom of choice, we first Skolemize to $\Sigma \overrightarrow{h}. \Pi \overrightarrow{a}. P$ and then transfer the variable $\overrightarrow{h}$ to the parameters. Concretely, the statement of surjectivity is skolemized to a surjection with an explicit right inverse, i.e.\ a split epi. We move the right inverse to the parameters. As a side effect, the distinction between classical and constructive mathematics virtually disappears for such explicit statements.

The addition of implicit Σ-types to Coq would change many things. Perhaps it would even allow us
to pack the proofs together with the operations, but it is too early to tell.
\end{comment}

@@

%Telescopes have been criticized~\cite{Packed} for the lack of multiple inheritance and
%the efficiency penalty of a long chain of coercion projections. Packed classes~\cite{Packed} provide
%a solution to these problems. We have provided an alternative solution based on type classes.

@@

% \section{Canonical structures}\label{canonical}
% Packed classes use canonical structures for the algebraic hierarchy. Both canonical structures and
% type classes may be seen as instances of hints in unification~\cite{Hints}. Some uses of canonical
% structures can be replaced by type class instances. The user manual (2.7.15) uses canonical
% structures to derive the setoid equality on the natural
% numbers in the following example \lstinline|(S x)==(S y)|. In this case type classes provide similar proof
% terms. Canonical structures give
% \begin{lstlisting}
% @equiv(Build_Setoid nat (@eq nat)(@eq_equivalence nat))(S x)(S y)
% \end{lstlisting}
% which includes an explicit proof that \lstinline|(@eq nat)| is an equivalence,
% whereas we obtain \lstinline|@equiv nat (@eq nat) (S x) (S y)|.

Unification hints are a more general mechanism than type classes, and could provide a more precise account of the interaction between implicit argument inference and proof search. It is not a great stretch to conjecture that a fruitful approach might be to use unification hints as the underlying mechanism, with type classes as an end-user interface encapsulating a particularly convenient idiom for using them.
% We encourage their inclusion to the Coq system. -- Eelis: not quite ready to go there yet

>>>>

@@

\begin{document}
\title{Type Classes for Mathematics in Type Theory}
\author{Bas Spitters \and Eelis van der Weegen} % For MSCS \\ Radboud University Nijmegen}
\institute{Radboud University Nijmegen}
\date{today}
\maketitle
\begin{abstract}
The introduction of first-class type classes in the Coq system calls for re-examination of the basic interfaces used for mathematical formalization in type theory.
We present a new set of type classes for mathematics and take full advantage of their unique features to make practical a particularly flexible approach formerly thought infeasible. Thus, we address both traditional proof engineering challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a minimum.

@@

In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transferred
to any type theory based proof assistant supporting type classes such as
Matita~\cite{asperti2007user}.

@@

@@

@@

Propositions in Coq are types~\cite{CMCP,ITT}, which themselves have types called \emph{sorts}. Coq features a distinguished sort called \lstinline|Prop| that one may choose to use as the sort for types representing propositions. The distinguishing feature of the \lstinline|Prop| sort is that terms of non-\lstinline|Prop| type may not depend on the values of inhabitants of \lstinline|Prop| types (that is, proof terms).
This regime of discrimination establishes a weak form of proof irrelevance, in that changing a proof can never affect the result of value computations. On a very practical level, this lets Coq safely erase all \lstinline|Prop| components when extracting certified programs to OCaml or Haskell.

@@

@@

@@

@@

@@

@@

@@

@@

All of \lstinline|Equiv|, \lstinline|RingPlus|, \lstinline|RingMult|, \lstinline|RingZero|, and \lstinline|RingOne| are operational (single-field) classes, with bound notations \lstinline|=|, \lstinline|+|, \lstinline|*|, \lstinline|0|, and \lstinline|1|, respectively.
\begin{wrapfigure}{r}{.4\textwidth}
    \includegraphics[scale=.8]{hierarchy.pdf}
\end{wrapfigure}
Let us briefly highlight some additional aspects of this style of structure definition in more detail.

@@

@@

@@

@@

@@

With these in place, our type class for categories follows the usual type-theoretical definition of a
category~\cite{saibi1995constructive}:
\begin{lstlisting}
  Class Category (O: Type) `{Arrows O} `{∀$$x y: O, Equiv (x ⟶ y)}
      `{CatId O} `{CatComp O}: Prop :=
    { arrow_equiv:> ∀ x y, Setoid (x ⟶ y)
    ; comp_proper:> ∀ x y z, Proper (equiv ==> equiv ==> equiv) comp
    ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
        c ◎$$ (b ◎$$ a) = (c ◎$$ b) ◎$$ a
    ; id_l `(a: x ⟶ y): cat_id ◎$$ a = a
    ; id_r `(a: x ⟶ y): a ◎$$ cat_id = a }.
\end{lstlisting}
This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.

@@

We introduce a convenient notation for the type of the computational content of a natural transformation between two functors:
\begin{lstlisting}
  Notation "F ⇛ G" := (∀ x, F x ⟶ G x).
\end{lstlisting}
Now assume the following context:
\begin{lstlisting}
  Context `{Category C} `{Category D}
    `{Functor (F: C → D)} `{Functor (G: D → C)}.
\end{lstlisting}
The naturality property is easy to write:
\begin{lstlisting}
  Class NaturalTransformation (η: F ⇛ G): Prop :=
    { naturaltrans_from: Functor F
    ; naturaltrans_to: Functor G
    ; natural: ∀ `(f: x ⟶ y), η$$ y ◎$$ fmap F f = fmap G f ◎$$ η$$ x }.
\end{lstlisting}

@@

@@

The rationals are characterized as a decidable field with an injective ring morphism from a canonical implementation of the integers and a surjection of fractions of such integers:
\begin{lstlisting}
  Context `{Field A} `{∀$$x y: A, Decision (x = y)} {inj_inv}.
  Class Rationals: Prop :=
    { rationals_field:> Field A
    ; rationals_frac: Surjective
        (λ p => integers_to_ring (Z nat) A (fst p) *
          / integers_to_ring (Z nat) A (snd p)) (inv:=inj_inv)
    ; rationals_embed_ints: Injective (integers_to_ring (Z nat) A) }.
\end{lstlisting}
Here, \lstinline|Z| is an \lstinline|Integers| implementation paramerized by a \lstinline|Naturals| implementation, for which we just take \lstinline|nat|. The choice of \lstinline|Z nat| here is immaterial; we could have picked another, or even a generic, implementation of \lstinline|Integers|, but doing so would provide no benefit.

@@

@@

For ease of presentation we show only a proof of concept for a very concrete language. We are currently working to integrate this technique with our existing universal algebra infrastructure. In particular, the latter's term data type should be ideally suited to serve as a generic symbolic representation of terms in a wide class of algebras. This should let us implement the basic setup of the technique once and for all, so that quotation for new algebraic structures can be enabled with minimal effort.

@@

Suppose now that we want to quote \lstinline|nat| expressions built from 1 and multiplication. To describe the relation we want the symbolic expression to have to the original expression, we first define how symbolic expressions evaluate to values (given a variable assignment):
\begin{lstlisting}
  Definition Value := nat.
  Definition Env V := V → Value.

@@

@@

@@

  Class Extend := extend: ∀ {x y} `{SemiGroupOp y} `{MonoidUnit y},
    (x → y) → (seq x → y).
  Class Singleton := singleton: ∀ x, x → seq x.
\end{lstlisting}
With these, we can define the predicate class for sequences:
\begin{lstlisting}
  Class Sequence
   `{∀ a, MonoidUnit (seq a)} `{∀ a, SemiGroupOp (seq a)}
   `{∀ a, Equiv a → Equiv (seq a)} `{Singleton} `{Extend}: Prop := ...
\end{lstlisting}
On top of this interface we can build theory about typical sequence operations such as maps, folds, their relation to \lstinline|singleton| and \lstinline|extend|, et cetera. We can also generically define `big operators' for sums ($\sum$) and products ($\prod$) of sequences, and easily show properties like distributivity, all without ever mentioning cons-lists.

@@

@@

@@

Unification hints are a more general mechanism than type classes, and could provide a more precise account of the interaction between implicit argument inference and proof search. It is not a great stretch to conjecture that a fruitful approach might be to use unification hints as the underlying mechanism, with type classes as an end-user interface encapsulating a particularly convenient idiom for using them.

There are really only two pending concerns that keeps us from making an unequivocal endorsement of type classes as a versatile, expressive, and elegant means of organizing proof developments. The first and lesser of the two is universe polymorphism for definitions as described in the previous section. The second is instance resolution efficiency. In more complex parts of our development we are now experiencing increasingly serious efficiency problems, despite having already made sacrifices by artificially inhibiting many natural class instances in order not to further strain instance resolution.
Fortunately, there is plenty of potential for improvement of the current instance resolution implementation. One source is the vast literature on efficient implementation of Prolog-style resolution, which the hint-based proof search used for instance resolution greatly resembles. We emphasize that these efficiency problems only affect type checking; efficiency of computation using type-checked terms is not affected.
